---
title: "Overall workflow for proteasome subunit analysis, Maria's paper"
output: html_document
date: "2025-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

######### libraries ##########
##############################
```{r}
## libraries
library(SingleCellExperiment)
library(Seurat)
library(tidyverse)
library(Matrix)
library(scales)
library(cowplot)
library(RCurl)
library(openxlsx)
library(knitr)
library(SeuratWrappers)
library(hdf5r)
library(reticulate)
library(data.table)
library(Matrix)
library(clustree)
library(data.table)
library(rstatix)
library(ggpubr)
```


### Point1:
"All proteasome subunits" expression comparing non-tumor, MB and GBM (Griesinger dataset from GlioVis). I have the plot done with only PSMB5, but we thought it might be good to have the same plot combining all proteasome subunits.

```{r}
## loadding the Griesinger dataset
pheno <- read.table("Griesinger/data/2026-03-09_Griesinger_pheno.txt", header = T )
expression <- fread("Griesinger/data/2026-03-09_Griesinger_expression.txt", header = T )

## Subsetting non-tumor samples, MB sample and GMB samples
pheno_sub<- pheno |> filter (Histology %in% c('Non-tumor', 'Medulloblastoma'))
expr_sub<- expression[expression$Sample %in% pheno_sub$Sample,]
str(expr_sub)

## Obtaining Thompson signature..
##### ##########################
proteasome_genes <- c(
  "PSMA1", "PSMA2", "PSMA3", "PSMA4", "PSMA5", "PSMA6", "PSMA7",
  "PSMB1", "PSMB2", "PSMB3", "PSMB4", "PSMB5", "PSMB6", "PSMB7", "PSMB8", "PSMB9",
  "PSMC1", "PSMC2", "PSMC3", "PSMC4", "PSMC5", "PSMC6",
  "PSMD1", "PSMD2", "PSMD3", "PSMD4", "PSMD5", "PSMD6", "PSMD7", "PSMD8", "PSMD9",
  "PSMD10", "PSMD11", "PSMD12", "PSMD13", "PSMD14",
  "PSME1", "PSME2", "PSME3") 

expr_proteasome <- expr_sub[, c("Sample", intersect(proteasome_genes, colnames(expr_sub))), with = FALSE] # 35/39
unique(colnames(expr_proteasome))

### ggplot of mean expression per group
expr_proteasome$histology<- pheno_sub$Histology
str(expr_proteasome)

proteasome_cols <- setdiff(colnames(expr_proteasome), c("Sample", "histology"))
expr_proteasome[, proteasome_mean := rowMeans(.SD, na.rm = TRUE), .SDcols = proteasome_cols]


ggplot(expr_proteasome, aes(x = histology, y = proteasome_mean)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  labs(
    x = "Histology",
    y = "Mean proteasome expression",
    title = "Proteasome mean expression by histology"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

### Two groups on this, so wilcoxox test 
# Do the mean proteasome expression scores differ between histology groups?
# Do group means differ after accounting for variability between patients?

unique(expr_proteasome$histology)
expr_proteasome$histology <- factor(expr_proteasome$histology,levels = c("Non-tumor", "Medulloblastoma"))
expr_proteasome[, .(
  n = .N,
  mean = mean(proteasome_mean, na.rm = TRUE),
  sd = sd(proteasome_mean, na.rm = TRUE),
  median = median(proteasome_mean, na.rm = TRUE),
  IQR = IQR(proteasome_mean, na.rm = TRUE)
), by = histology]

expr_proteasome$histology <- factor(
  expr_proteasome$histology,
  levels = c("Non-tumor", "Medulloblastoma")
)

#stat.test <- aov(proteasome_mean ~ histology, data = expr_proteasome) %>% tukey_hsd() # ANOVA, tukey posthoc

## Include results of ANOVA posthoc in the graph 
p<- ggpubr::ggviolin(expr_proteasome, x = "histology", y = "proteasome_mean",
                 palette = "jco", color = "histology", add = "jitter", legend = "right") +
  labs(
    x = NULL,
    y = "Mean Log2FC expression"
  ) +
  theme(
    axis.text.x = element_text(size = 14 ),
    axis.text.y = element_text(size = 14 ),
    axis.title.y = element_text(size = 16),
    axis.title.x = element_blank()
  ) + RotatedAxis()

pdf("Griesinger/Mean_proteasome_exp.pdf", width = 7, height = 5)
p + stat_compare_means()
dev.off()

#write.csv(stat.test,  "Griesinger/ANOVA_wilcoxox_stats.csv")
write.csv(expr_proteasome,  "Griesinger/proteasome_expression.csv")
```


### Point2:
"All proteasome subunits" expression comparing MB subgroups (I was using Northcott dataset to do the plot for PSMB5 expression level).
```{r}
## loadding the Griesinger dataset
pheno <- read.table("Cavalli/data/2026-03-09_Cavalli_pheno.txt", header = T )
expression <- fread("Cavalli/data/2026-03-09_Cavalli_expression.txt", header = T )

## Subsetting non-tumor samples, MB sample and GMB samples
all(pheno$Sample == expression$Sample)
expression$Subgroup<- pheno$Subgroup

## Obtaining Thompson signature..
##### ##########################
proteasome_genes <- c(
  "PSMA1", "PSMA2", "PSMA3", "PSMA4", "PSMA5", "PSMA6", "PSMA7",
  "PSMB1", "PSMB2", "PSMB3", "PSMB4", "PSMB5", "PSMB6", "PSMB7", "PSMB8", "PSMB9",
  "PSMC1", "PSMC2", "PSMC3", "PSMC4", "PSMC5", "PSMC6",
  "PSMD1", "PSMD2", "PSMD3", "PSMD4", "PSMD5", "PSMD6", "PSMD7", "PSMD8", "PSMD9",
  "PSMD10", "PSMD11", "PSMD12", "PSMD13", "PSMD14",
  "PSME1", "PSME2", "PSME3") 


expr_proteasome <- expression[, c("Sample", intersect(proteasome_genes, colnames(expression))), with = FALSE] #39/39
unique(colnames(expr_proteasome))

### ggplot of mean expression per group
all(pheno$Sample == expr_proteasome$Sample)
colnames(expr_proteasome)
expr_proteasome$Subgroup<- pheno$Subgroup
str(expr_proteasome)

proteasome_cols <- setdiff(colnames(expr_proteasome), c("Sample", "Subgroup"))
expr_proteasome[, proteasome_mean := rowMeans(.SD, na.rm = TRUE), .SDcols = proteasome_cols]


ggplot(expr_proteasome, aes(x = Subgroup, y = proteasome_mean)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  labs(
    x = "Histology",
    y = "Mean proteasome expression",
    title = "Proteasome mean expression by Subgroup"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

### Anova test with Tukey posthoc correction
# Do the mean proteasome expression scores differ between histology groups?
# Do group means differ after accounting for variability between patients?

unique(expr_proteasome$Subgroup)
expr_proteasome[, .(
  n = .N,
  mean = mean(proteasome_mean, na.rm = TRUE),
  sd = sd(proteasome_mean, na.rm = TRUE),
  median = median(proteasome_mean, na.rm = TRUE),
  IQR = IQR(proteasome_mean, na.rm = TRUE)
), by = Subgroup]

unique(expr_proteasome$Subgroup)
expr_proteasome$Subgroup <- factor(
  expr_proteasome$Subgroup,
  levels = c("SHH" ,"Group3", "Group4", "WNT")
)

stat.test <- aov(proteasome_mean ~ Subgroup, data = expr_proteasome) %>% tukey_hsd() # ANOVA, tukey posthoc

## Include results of ANOVA posthoc in the graph 
pdf("Cavalli/Mean_proteasome_exp.pdf", width = 7, height = 5)
ggpubr::ggviolin(expr_proteasome, x = "Subgroup", y = "proteasome_mean",
                 palette = "jco", color = "Subgroup", add = "jitter", legend = "right") +
  stat_pvalue_manual(
    stat.test, label = "p.adj.signif", 
    y.position = c(9.5,9.7,9.9,10.1,10.3,10.5)
      ) +
  labs(
    x = NULL,
    y = "Mean LogFC proteasome expression"
  ) +
  theme(
    axis.text.x = element_text(size = 14 ),
    axis.text.y = element_text(size = 14 ),
    axis.title.y = element_text(size = 16),
    axis.title.x = element_blank()
  ) + RotatedAxis()
dev.off()

write.csv(stat.test,  "Cavalli/ANOVA_tukey_posthoc_stats.csv")
write.csv(expr_proteasome,  "Cavalli/proteasome_expression.csv")


expr_proteasome[, .(
  n = .N,
  mean = mean(proteasome_mean, na.rm = TRUE),
  sd = sd(proteasome_mean, na.rm = TRUE),
  median = median(proteasome_mean, na.rm = TRUE)
), by = Subgroup][order(-mean)]
```












